---
title: "Broccoli GWAS of Horticultural Quality Traits"
author: "Zachary Stansell, Sean O'Neill, Thomas Björkman"
date: '`r format(Sys.Date(), "%d-%b-%Y")`'
output:
  html_document:
    highlight: pygments
    theme: united
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
link-citations: true
always_allow_html: true
---

# Setup
## R setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
source("bin/R/PACKAGES.R") 
#library("tidyverse")
```
##bash setup
```{bash, eval=FALSE}

#add code here for paths, dependencies, etc
bin/shell/INSTALL.sh" 


```



# Introduction and Learning 

## Standing Questions

* What's the proper application of nesting (in an experimental design: 2 years, 2 trials/year, 3 reps) **answer**  
* Should all 4 experiments be analyzed independent or bulked? **answer**  
* Should multilevel models be used? **probably not**  
* Should year/trial/rep/genotype be factors (tentative yes) **answer**  
* What are the dependent and independent variable in this analysis. **answer**  
* Is it fair to use GWAS with categorical data (e.g., flower color) **no**  
* What types of data other than continuous are allowed? how are they typically treated? **answer**  
* Does our sample size (N=97) make sense? what are similar GWAS studies that have used small sample size? **answer**  
* Is there an issue using REP (compared to year and trial) because... ... ... ... it doesn't .... ....... ... ... ... 
* Can you show that FOR ALL TRAITS, rep within trait isn't important? 
	* For now, lets ASSUME that there's no significant rep effect...
* Why would you normalize data for GWAS? **answer**

**Notes from Feb 2, 2021**

* Population structure-- options available, look at variation autoencoder resolving pop structure more clearly? (currently using TASSEL for PCA and fastSTRUCTURE for population structure-- what about structure using variations autoencoders? what happens?)  
* Normalization--- read about, 




## Data Exploration

```{r data_exploration_and_learning, eval=FALSE, echo=FALSE, warning=FALSE  }
# spend time testing the data.. are there any entry errors?
# are there outliers?
# make tables e.g., table(pheno$sc) or table(pheno$FC, pheno$geno)
# make histograms to look for outliers; hist(pheno$BM)

pheno <-
	read.csv(file = "raw/PHENO/raw/pheno_final.csv",
					 stringsAsFactors = FALSE,
					 na.strings = "NA") 


# Compare trial years by mean values
pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4)) %>%
	group_by(geno, trial) %>%
	summarize(mean = mean(F50, na.rm = TRUE), .groups = "drop") %>%
	spread(trial, mean) %>%
	select(-geno) %>%
	pairs()



# select only year 2
table(test$geno[test$year == 2])

# use tidy to select year two

pheno %>% dplyr::filter(BS > 35)



subset_data <-
	pheno[, c(
		"MT",
		"FF",
		"F50",
		"MF50",
		"HT",
		"WD",
		"BM",
		"LT",
		"HM",
		"HI",
		"HD",
		"CL",
		"BS",
		"BR",
		"BU",
		"HC",
		"HE",
		"HS",
		"HU",
		"HR",
		"OQ",
		"FC"
	)]

pairs(subset_data)

boxplot(pheno$MT ~ pheno$trial)


traits <-c("MT","FF","F50","MF50","HT","WD","BM","LT","HM","HI","HD","CL","BS","BR","BU","HC","HE","HS","HU","HR","OQ","FC")



pdf("rst/sandbox/shittyboxplot.pdf", height= 20, width=20)
par(mfcol=c(6,4))
for (i in traits){
	boxplot(pheno[,i] ~ pheno$trial, main= i)
}
dev.off()

# Compare trial years
#throw out "trial 0"
pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4)) %>%
	group_by(geno, trial) %>%
	summarize(mean = mean(BR, na.rm = TRUE), .groups = "drop") %>%
	spread(trial, mean) %>%
	select(-geno) %>%
	pairs()
	
# 
temp <- pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4))

lmMod <- lm(MT ~ geno, data=temp)

# initial model
#Now that the model is ready, there are two ways to test for heteroscedasticity:
#Graphically Through statistical tests
#Graphical method

par(mfrow=c(2,2)) 
# init 4 charts in 1 panel
plot(lmMod)



lmMod <- pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4)) %>%
	lm(. ,HI ~ geno) # initial model
#Now that the model is ready, there are two ways to test for heteroscedasticity:

#Graphically
#Through statistical tests
#Graphical method
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(lmMod)

# Example of coloring by trial
pairs(temp[,10:18], pch = 21,bg = c("red", "green3", "blue", "white", "black")[temp$trial])


## Stuff we've learned
* When comparing 2019 trials (1 and 2) with 2020 trials (3 and 4).
	* 2020 was wider, heavier, a bit earlier. overall, larger in every way.
	* in fact, head mass was 86% larger and above ground biomass was 96% larger!
	* There seems to be more laterals (in which year? year 2?)
	
	
## questions

* How do we aggregate the data?
* Do we summarize by year?


## ANOVA,how to aggregate, heteroscedasticity
*figure out how to pair between trials, reps
*aggregation possibilities ()
	*average of reps with each planting
	*by planting or year
		*average 1 and 2, 3 and 4
		*average 1 and 3, 2 and 4
*q: which aggregations are appropriate?
	*ANOVA
	*Look at weather data, population structure.  try to reveal error structure (?)
	*see how different aggregations affect error, heteroscedasticity
*q: Which variables should be studied in GWAS
	*read GWAS to understand conventions, limitations, what makes traits good candidates 
*q: should any accessions be filtered?

	
```

```{r pair_by_trial, eval = FALSE}
# a way of plotting pairs by trial.  note that the number of accessions differs by trial
# consider writing a function to generalize

trial.sizes <- table(temp$trial);
big.trial <- which(trial.sizes == max(trial.sizes));


```

## Learning ANOVA
	

```{r models_learning, eval=FALSE}


######################
# here we're inquiring about rep effects.
# load data
pheno <- read.csv("raw/PHENO/raw/pheno_final.csv") %>% dplyr::filter(trial != 0)


# question.. are the reps different within planting?
par(mfcol = c(2, 2))

for (i in levels(as.factor(pheno$trial))) {
	test <- pheno %>% dplyr::filter(trial == i)
	boxplot(test$WD ~ as.factor(test$rep), main = paste("trial",i), ylim=c(0,200))
	
}


summary(
	aov(pheno$MT ~ as.factor(pheno$trial) + as.factor(pheno$rep))
	)

# For MT (days to maturity) we 'fail to reject the null (H0) that there's no diff between reps)


######################



# select only year 2
table(test$geno[test$year == 2])

# use tidy to select weird bead sizes:
pheno %>% 	dplyr::filter(BS > 35)


subset_data <-
	pheno[, c(
		"MT",
		"FF",
		"F50",
		"MF50",
		"HT",
		"WD",
		"BM",
		"LT",
		"HM",
		"HI",
		"HD",
		"CL",
		"BS",
		"BR",
		"BU",
		"HC",
		"HE",
		"HS",
		"HU",
		"HR",
		"OQ",
		"FC"
	)]
pairs(subset_data)

boxplot(pheno$MT ~ pheno$trial)


traits <-
	c(
		"MT",
		"FF",
		"F50",
		"MF50",
		"HT",
		"WD",
		"BM",
		"LT",
		"HM",
		"HI",
		"HD",
		"CL",
		"BS",
		"BR",
		"BU",
		"HC",
		"HE",
		"HS",
		"HU",
		"HR",
		"OQ",
		"FC"
	)


pdf("rst/sandbox/shittyboxplot.pdf", height= 20, width=20)
par(mfcol=c(6,4))
for (i in traits){
	boxplot(pheno[,i] ~ pheno$trial, main= i)
}
dev.off()

# Compare trial years
#throw out "trial 0"
pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4)) %>%
	group_by(geno, trial) %>%
	summarize(mean = mean(BR, na.rm = TRUE), .groups = "drop") %>%
	spread(trial, mean) %>%
	select(-geno) %>%
	pairs()
	
# 
temp <- pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4))

# initial model
#Now that the model is ready, there are two ways to test for heterosedasticity:
#Graphically Through statistical tests
#Graphical method

par(mfrow=c(2,2)) 
# init 4 charts in 1 panel
plot(lmMod)



lmMod <- pheno %>%
	dplyr::filter(trial %in% c(1, 2, 3, 4)) %>%
	lm(. ,HI ~ geno) # initial model
#Now that the model is ready, there are two ways to test for heterosedasticity:

#Graphically
#Through statistical tests
#Graphical method
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(lmMod)

#Example of coloring by trial
pairs(temp[,10:18], pch = 21,bg = c("red", "green3", "blue", "white", "black")[temp$trial])


# converted year, trial, geno, rep of temp to factors.  was this a good idea
# spend time learning syntax of formulae
# installed 
# install.packages("lmerTest")
library(lmerTest)

# aov example.
HT.aov <- aov(HT ~ year:trial, temp)

# apply to all traits, summarize 

# messing with multilevel

#HT.lmer <- lmer(HT ~ geno + (1 | year:trial), data=temp)
#HT.lmer <- lmer(HT ~ geno + (1 | year:trial:rep), data=temp)
#HT.lmer <- lmer(HT ~ geno + (rep | year:trial), data=temp) #figure out what random slope means and why this gives an error.

```





# Phenotype

## Cleaning data

```{r  Cleaning Data, eval=FALSE}
source("bin/R/cleaningPhenoData.R") 

```


# SNP Calling


```{bash, eval=FALSE}

#see notes inside /bin/shell/GB


# Configuration. LOTS OF REDUNDANCY UNTIL THIS MAKES SENSE



##Edit default settings here.

###make a note of where we are.  is this silly?
HERE=$PWD

### Set working directory for GBS and FILTERING_SNPS
DIR="/home/$USER/PROJECTS/broccoli_gwas/raw/GBS/"

### create a unique name so that files are not overwritten. For now we will list names of directories in REF_GENOME to iterate through reference genomes
cd $DIR"REF_GENOME"
NAME=*
cd $HERE

### maximum RAM for TASSEL to use.  e.g. "-Xmx25g" for 25gb of RAM.  (ADD SETTINGS FOR MIN AND MAX????)
MEM="-Xmx25g"

### specify the path to the key
KEY=$DIR"KEY/masterKey.txt"

### specify where the reference genome(s) is(are) located. 
REFGENOME=$DIR"REF_GENOME/Brassica_oleracea.BOL.dna.toplevel.fa.gz"

### Set directory with Illumina raw files.
ILLUMINA=$DIR"ILLUMINA"

### set where the tassel db is located
DB=$DIR

### where to output results/summary stats from pipeline
STATS=$DIR  

### path to find the taxa list
TAXA=$DIR"N109.txt"

###BOL directory
BOL=$DIR"BOL/"

### where to output final h5 file procuded.
OUTFILE=$DIR

for i in $NAME
  do
    echo i
    #bin/shell/GBS.sh --DIR=$DIR --NAME=$NAME --MEM=$MEM --KEY=$KEY --REFGENOME=$DIR"REF_GENOME"$i"/"* --ILLUMINA=$ILLUMINA --DB=$DB --STATS=$STATS --OUTFILE=$OUTFILE

    #bin/shell/FILTERING_SNPS.sh --DIR=$DIR --NAME=$NAME --MEM=$MEM --TAXA=$TAXA --BOL=$BOL --OUTFILE=$OUTFILE
  done

#GOAL IS TO MAKE A FLEXIBLE PIPELINE THAT PRODUCES A FILTERED .h5 FILE NO MATTER WHAT WE FEED IT

```






# GWAS

## Transforming Data

```{r GWAS, eval=FALSE}
source("bin/R/normalizingFun.R")

PSP <- read.csv("raw/PHENO/raw/PSP.csv")


gwasTraits <- c(
		"gbs",
    "HT",
    "WD",
    "HE",
    "LT",
    "BM",
    "HM",
    "HI",
    "BS",
    "BU",
    "HC",
    "HD",
    "CL",
    #"RR",
    "BR",
    "HS",
    "HU",
    "HR",
    "OQ",
    # note, when only using F1, all flower color is the same
    #"FC", 
    "MT",
    "FF",
    "F50",
    "MF50" #,
    # "PC1",
    # "PC2",
    # "PC3"
  )


# Normalize ALL data
# gwasNN.ALL == gwas phenotypes, not normalized, all taxon
gwasNN.ALL <-  PSP  %>% 
  #filter(Type %in% "F1") %>% 
  dplyr::select(gwasTraits)




#Make a table appropriate for TASSEL GUI:
gwasNN.ALL <-
	rbind(c("taxa", rep("data", dim(gwasNN.ALL)[2] - 1)),
				colnames(gwasNN.ALL),
				as.matrix(gwasNN.ALL))

colnames(gwasNN.ALL) <-
 	c("<Phenotype>", rep("", dim(gwasNN.ALL)[2] - 1))

write.table(
	gwasNN.ALL,
	"raw/PHENO/normalization/gwasNN.ALL.txt",
	quote = FALSE,
	row.names = FALSE,
	sep = "\t"
)




# Normalize data using power functions


n1 <- lapply(gwasNN.ALL[-1],normalizingFun)
n2 <- as.data.frame(sapply(n1, `[`))


gwasNorm.ALL <- cbind(gwasNN.ALL$gbs, n2)
colnames(gwasNorm.ALL)[1] <- "gbs"

# write data
write.table(
	gwasNorm.ALL,
	"../OUTPUT/GWAS/pheno_gwas_ALL.txt",
	sep = "\t" ,
	quote = FALSE,
	row.names = FALSE
)

# Normalize F1 data   
gwasNN.F1 <-  PSP  %>% 
  filter(subpop %in% "Calabrese.F1") %>% 
  dplyr::select(gwasTraits)

n1 <- lapply(gwasNN.F1[-1],normalizingFun)
n2 <- as.data.frame(sapply(n1, `[`))
gwasNorm.F1 <- cbind(gwasNN.F1$gbs, n2)
colnames(gwasNorm.F1)[1] <- "gbs"

# write data
write.table(gwasNorm.F1,"../OUTPUT/GWAS/pheno_gwas_F1.txt", sep = "\t" , quote=FALSE, row.names = FALSE)

# Normalize LR data
gwasNN.LR <-  PSP  %>% 
  filter(subpop %in% c("Calabrese.LR","Sprouting.Broccoli","Violet.Caul")) %>% 
  dplyr::select(gwasTraits)

n1 <- lapply(gwasNN.LR[-1],normalizingFun)
n2 <- as.data.frame(sapply(n1, `[`))
gwasNorm.LR <- cbind(gwasNN.LR$gbs, n2)
colnames(gwasNorm.LR)[1] <- "gbs"

# write data
write.table(gwasNorm.LR,"../OUTPUT/GWAS/pheno_gwas_LR.txt", sep = "\t" , quote=FALSE, row.names = FALSE)


#QQ plots

gwasListNN <-
	list(gwasNN.ALL = gwasNN.ALL,
			 gwasNN.F1 = gwasNN.F1,
			 gwasNN.LR = gwasNN.LR)


gwasListNorm <-
	list(
		gwasNorm.ALL = gwasNorm.ALL,
		gwasNorm.F1 = gwasNorm.F1,
		gwasNorm.LR = gwasNorm.LR
	)

#choose data
for (j in 1:3) {
  gwasNN <- as.data.frame(gwasListNN[j])
  gwasNorm <- as.data.frame(gwasListNorm[j])
# make plots loop  
 pdf(
    paste("../OUTPUT/GWAS/qqPlot/", names(gwasNN)[j], ".pdf"),
    height = 50,
    width = 6
  )
  par(mfrow = c(22, 2), mar = c(1.75, 2, 1.4, 1))
  
  
  for (i in 1:25) {
    #qq of non-normalized results
    qqPlot(
      gwasNN[, i + 1],
      main = colnames(gwasNN[i + 1]),
      xlab = "",
      ylab = "" ,
      col = alpha(viridis(7)[j], .75),
      pch = 19,
      cex = 1,
      col.lines = alpha(viridis(7)[j], 0.5),
      id = FALSE
    )
    text(-1.75, max(gwasNN[, i + 1], na.rm = TRUE),
         paste("p = ", round(shapiro.test(gwasNN[, i + 1])$p.value, 3)), font =
           2)
    
    #qq of normalized results
    qqPlot(
      gwasNorm[, i + 1],
      main = colnames(gwasNN[i + 1]),
      xlab = "",
      ylab = "" ,
      col = alpha(viridis(7)[j*2], .75),
      pch = 19,
      cex = 1,
      col.lines = alpha(viridis(7)[j*2], 0.5),
      id = FALSE
    )
    text(-1.75,
         max(gwasNorm[, i + 1], na.rm = TRUE),
         paste("p = ", round(shapiro.test(gwasNorm[, i + 1])$p.value, 3)),
         font = 2)
    LAMDA <- powerTransform(gwasNN[, i + 1], family = "yjPower")
    text(
      -1.75,
      0.95 * max(gwasNorm[, i + 1], na.rm = TRUE),
      ifelse(
        shapiro.test(gwasNN[, i + 1])$p.value > 0.05,
        "dist is normal",
        paste("lamda = ", round(LAMDA$lambda, 2))
      )
    )
  }
  dev.off()
}


# Make GWAS phenotype data, eval=FALSE}

pheno <- read.table("../OUTPUT/GWAS/pheno_gwas_ALL.txt", head = TRUE )
columbz <-c("taxa",rep("data",dim(pheno)[2]-1))
test <- rbind(columbz,colnames(pheno), as.matrix(pheno))
colnames(test) <- c("<Phenotype>",rep("",dim(test)[2]-1))
write.table(test,"../OUTPUT/GWAS/pheno_gwas_normalized_ALL.txt", quote=FALSE, row.names=FALSE, sep="\t" )

pheno <- read.table("../OUTPUT/GWAS/pheno_gwas_F1.txt", head = TRUE )
columbz <-c("taxa",rep("data",dim(pheno)[2]-1))
test <- rbind(columbz,colnames(pheno), as.matrix(pheno))
colnames(test) <- c("<Phenotype>",rep("",dim(test)[2]-1))
write.table(test,"../OUTPUT/GWAS/pheno_gwas_normalized_F1.txt", quote=FALSE, row.names=FALSE, sep="\t" )

pheno <- read.table("../OUTPUT/GWAS/pheno_gwas_LR.txt", head = TRUE )
columbz <-c("taxa",rep("data",dim(pheno)[2]-1))
test <- rbind(columbz,colnames(pheno), as.matrix(pheno))
colnames(test) <- c("<Phenotype>",rep("",dim(test)[2]-1))
write.table(test,"../OUTPUT/GWAS/pheno_gwas_normalized_LR.txt", quote=FALSE, row.names=FALSE, sep="\t" )

rm(list=ls())
```



## Run Tassel
```{bash TASSEL GLM, eval=FALSE}

bin/shell/GWAS.sh

```




## Plot Data
```{r plot GWAS results, eval=FALSE}
source("bin/R/manhattan.R")

GLM <-
	read.delim("rst/sandbox/GLM_ALL1.txt",
						 sep = "\t",
						 na.strings = "NaN")

GLM_PCA <-
	read.delim("rst/sandbox/GLM_ALL_PCA1.txt",
						 sep = "\t",
						 na.strings = "NaN")


# MLM_PCA_K <- read.delim("../OUTPUT/GWAS/CLI/MLM_PCA_K23.txt", sep="\t", na.strings = "NaN")
# names(MLM_PCA_K)[names(MLM_PCA_K) == "Locus"] <- "Chr"
# MLM_Q_K <- read.delim("../OUTPUT/GWAS/CLI/MLM_Q_K23.txt", sep="\t", na.strings = "NaN")



# 
# names(MLM_Q_K)[names(MLM_Q_K) == "Locus"] <- "Chr"
# #temp fix for some p values == 0? idk why
# MLM_Q_K <- na.omit(MLM_Q_K)
# MLM_Q_K[MLM_Q_K$p == 0,] <- 1e-16


plotGWAS(GLM,  filename="rst/sandbox/GLM.png", res=300)
plotGWAS(GLM_PCA,  filename="rst/sandbox/GLM_PCA.png", res=300)




# plotGWAS(MLM_PCA_K,  filename="../OUTPUT/GWAS/MLM_PCA_K.png", res=150)
# plotGWAS(MLM_Q_K,  filename="../OUTPUT/GWAS/MLM_Q_K.png", res=150)
```
